Ant cuticle image classification using texture analysis: a comparative study

### 2022-07-08 paper/main.tex
add kviews parameters
address comments first

\begin{abstract}
    There is a large variety of ant species, and most species are diverse in
    terms of size, shape, behaviors, and especially cuticle textures. However,
    the significance of ant cuticle texture is not widely researched. Ant
    cuticle texture presumably provides some type of function, and therefore is
    useful to research for ecological applications and bioinspired designs. This
    research employs image texture analysis and deep machine learning to
    automatically group similar ant species based on morphological traits. We
    provide a comparative study of the performance of general image
    classification and texture analysis methods on ant head images. We evaluate
    the results of the classification methods with modern visualization
    techniques. Our results show that deep learning methods are applicable for
    classifying ant cuticle textures. We also show that the performance of the
    deep texture analysis methods performs better than the deep classification
    methods.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Research has heavily focused on the macrostructures
and internal chemical components that make the exoskeleton functional and more
recent work is being done to understand the functional aspects of external
cuticle micro sculpturing \cite{muthukrishnan_insect_2020,
    gunderson_insect_1989, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires specialized expertise.
Therefore, automated entomology is gaining attraction by both biologists and
computer scientists and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can take advantage of existing work in general
image analysis methods.

@@

Our primary contribution is the custom dataset used to explore the relationship
between ant cuticle texture and automated texture analysis methods. We compare
classical and modern image classification methods and texture analysis methods
by evaluating their performance on our dataset.

\section{Related work and literature review}

In this section we will review the literature for ant sculpturing identification
and texture analysis. Additionally, we discuss the work related to automated
insect identification.

\subsection{Sculpturing identification}
Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}.

@@

There is often substantial overlap among terms, and closely related cuticle
patterns are likely to be functionally similar. For example, the definition for
“imbricate” is, “partly overlapping and appearing like shingles on a roof or
scales on a fish,” which is difficult to distinguish from the definition of
tesselate, “made up of squares like a chess board, either in sculpturing or in
color.” However, we note that the terms used to describe the cuticle patterns
are similar to the terms used to describe textural patterns in texture analysis.
In the next subsection, we review some literature of image texture analysis.

\subsection{Texture analysis}

Image texture analysis is an important part of image interpretation as many
objects appearing in the textural patterns either partially or completely inside
an image. This phenomenon appears frequently in many images we encountered in
our environment as almost everybody is using the iPhone to capture the photos.
Scientists are also studying their subjects depending on taking the images
remotely such as remotely sensed satellite images. With an abundant number of
textural images surrounding us, it is essential to have a reliable automatic
tool for the textural image interpretation.

Classical computer methods are developed for textural images based on the
spatial relationship of gray levels of pixels encoded in the texture
\cite{haralick_image_1985,hung_image_2019}. Those methods by de-facto follow the
steps used in the statistical pattern recognition
\cite{fukunaga_introduction_2013}. Therefore, textural features are extracted,
and the classification and segmentation methods are used for the interpretation
either using the supervised or unsupervised approaches. The interpretation
method is usually taken and modified from the traditional pattern recognition
methods. For example, the gray-level co-occurrence matrix (GLCM) and local
binary patterns (LBP) are two of methods for extracting features
\cite{haralick_textural_1973, goos_gray_2000}.

Similar to the traditional pattern recognition, classical image texture analysis
methods can be grouped into four categories: statistical, structural,
model-based, and transform-based methods \cite{bharati_image_2004}. Among all of
the categories, the statistical method is often used by employing the method
from the pattern recognition principle.  In doing so, an image texture is
represented as feature vectors and then fed into the algorithm. Markov random
field models are also studied for textural image interpretation
\cite{hassner_use_1981, cross_markov_1983}. The transform-based methods use some functions to
decompose an image texture into a set of basic feature images. Gabor filters and
Wavelet expansions are two of the widely used approaches
\cite{bovik_multichannel_1990}.

@@

Other network architectures for end-to-end texture learning are also available.
For example, in the deep multiple-attribute-perceived network (MAP-Net)
\cite{zhai_deep_2019}, multiple perceptual attributes are progressively learned
in a mutually reinforced manner through multiple branches. In the deep
structure-revealed network (DSR-Net) \cite{zhai_deep_2020}, inherent structural
representation for a texture pattern is obtained by employing a primitive
capturing module to learn spatial primitives and a dependency learning module to
capture the dependency among the primitives. In \cite{mao_deep_2021}, a residual
pooling layer consisting of a residual encoding module and an aggregation module
is used to generate discriminative features of low dimensions. In
\cite{peeples_histogram_2021}, a histogram layer is designed to compute local
spatial distribution of CNN features. In \cite{chen_deep_2021}, an innovative
aggregation module is presented to exploit statistical self-similarity across
layers. All these architectures customize the standard CNN structure to
accomplish the characterization of certain spatial, visual, or statistical
nature unique to textural images.

@@

\subsection{Insect classification}
In this section, we provide an overview of some insect classification methods.
Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, \textit{e.g.} \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, \textit{e.g.} \textit{ant}, \textit{fly}, and \textit{bee}
\cite{deng_imagenet_2009}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

\section{Methodology}

In this section, we provide an overview of the methodology used in this paper.
We describe the steps used to prepare the custom dataset. Finally, we describe
the algorithms applied and experimental setup to obtain the results.

\subsection{Dataset preparation} % preparation generation

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/rough_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:rough-cuticle-texture}
\end{figure}

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. We also
refer to these ant head images, which describe the pose of the ant, as ant
cuticule images, due to the clear cuticle texture present on the head. Some
randomly selected images from each category are shown in Figures
\ref{fig:rough-cuticle-texture} and \ref{fig:smooth-cuticle-texture}.

@@

To begin, a master spreadsheet was created with the 2,499 different ant species
to be identified for the primary dataset. The team was trained to identify
cuticle sculpturing through a process which consisted of one 45-minute
introductory lesson explaining the project and texture categories. Then, the
team was given a training set of photos to identify from the genus
\textit{Polyrhachis}. The sculpture identification protocol describes the two
primary categories: \textit{rough} and \textit{smooth}.

@@

To collect the images, the assistants followed the taxonomy information
available in the master spreadsheet to the appropriate AntWeb page
\cite{perrichot_antweb_2012}. In many cases, there are multiple ant head images
of the same species, and occasionally there are multiple image resolution
available from a single image. To simplify the data collection process, the
assistants were instructed to download the first ant head image of the species
being identified in the highest resolution possible. Each image was named with
an identifier that corresponds with the row number in the master spreadsheet.
The same ant head images that were downloaded in the data collection phase were
the same ones used in the sculpture identification protocol. Ant species which
did not have any images of the head were excluded from the dataset.
Additionally, ant species which only had a head image of a queen ant were
excluded from the dataset.

@@

Due to the variety of the ant head image attributes, we apply simple
preprocessing before the images are used in our model. We want the images to
have a uniform size for simplicity in our classification process. Since the ant
head images are typically centered in the image, we apply a center crop to each
image to create a square image of the same size. Once the image is square, we
resize each image to a fixed size of 256x256 pixels. We leave other
discrepancies in the images untouched.

\subsection{Classical texture analysis methods}

@@

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 10-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm.

\subsection{Deep learning models used for the experiments}

Our first model is \textit{visual geometry group} (VGG), a convolutional neural
network that takes advantage of very small convolutional filters in a deep
network architecture \cite{simonyan_very_2015}. We compare four architectures of
VGG: VGG11, VGG13, VGG16, and VGG19. The primary difference between the
architectures is the number of layers in each model. Our second model is
\textit{residual network} (ResNet), a deep network architecture that includes
shortcut connections between layers (residual connections) \cite{he_deep_2015}.
We compare three architectures of ResNet: ResNet18, ResNet50, and ResNet101.
Again, the primary difference between the architectures is the number of layers
in each model.

\subsection{Deep texture analysis method used for the experiments}

@@

\section{Results and analysis}
\subsection{Environment}

@@

\subsection{Evaluation}

\subsection{Experimental results}

For the experiments using the deep learning algorithms, we use our custom
dataset of ant head images which contains 2,499 images. 1072 samples of rough
textured ant cuticle textures comprise 43\% of the dataset. The remaining 1427
samples of smooth textured ant cuticle textures comprise 57\% of the dataset. To
handle the imbalance of the dataset, we apply undersampling for each class for
the training dataset. By using random stratified sampling, we construct a
training set with 800 images per class. The remaining images are randomly split
between test and validation, which turns out to roughly a 60\%/20\%/20\% train,
test, and validation data split. With 272 rough samples and 627 smooth samples
left over after the stratified split, the test dataset has roughly 136 rough
samples and 313 smooth samples. Since these leftover samples are split with code
by 50\% there will be some rounding variance and therefore the test dataset
built at run-time will not always have exactly the same number of samples.

To begin, we share the results on the each algorithm specified in the
methodology section using the evaluation metrics described. The results are
shown in Table \ref{tab:results}. It should be noted that due to the class
imbalance in the dataset, the F1 score is the preferable metric to the accuracy.
Therefore, the results have been sorted by F1 score in descending order.

@@

\subsection{t-SNE Visualization}

@@

In this section, we provide visualization of the fine-tuned ResNet101 model and
the randomly initialized ResNet101 model using t-SNE dimensionality reduction.
First, we run the dataset preprocessing method and initialize both models. Then,
both models are trained according to the training parameters and the state of
each model is saved. To visualize the deep extracted features, we modify each
model to obtain the embeddings of the second to last layer. Then, we use the
t-SNE algorithm to reduce the dimensionality of the embeddings to 2 dimensions.
We plot side-by-side the ground truth and predicted labels for each model.
Figure \ref{fig:resnet101_tsne} shows the results of the trained randomly
initialized model and Figure \ref{fig:fresnet101_tsne} shows the results of the
fine-tuned model.

@@

\FloatBarrier
Based on the visual results of the t-SNE visualization, we can see that the
fine-tuned model learned a stronger separation of the two classes, which
reinforces the results that the fine-tuned model received a higher average
accuracy.

\FloatBarrier
\subsection{GradCAM visualization}

In this section, we provide some visual analysis of some correctly and
incorrectly classified images using GradCAM. We provide two categories and two
subcategories in our analysis. The two categories are correct and incorrect
classification. Regardless of the feature activation map, the correctly
classified images have the same predicted label as the ground truth, and
incorrectly classified images have different predicted labels. The two
subcategories are expected and non-expected feature activation. In the expected
case, the features that are used to compute the classification are the same as
the features used by the assistants in the sculpture identification process. In
general, the features used by the assistants are the textures of the cuticle on
the ant head. In the non-expected case, the features used to compute the
classification are not from the head, for example, from the background,
extraneous text, or the body of the ant. We used randomly selected images from
the dataset and the fine-tuned ResNet101 model to perform the analysis. We show
the GradCAM results in Figures \ref{fig:correct_ideal},
\ref{fig:correct_nonideal}, \ref{fig:incorrect_ideal}, and
\ref{fig:incorrect_nonideal}. The left image shows the preprocessed image input
to the model. The right image shows the GradCAM output based on the
classification. Four specimens were selected randomly from each category and
subcategory.

Correctly classified images which use the expected features show the ideal
performance of the model. Incorrectly classified images which use the expected
features should be further analyzed. In essence, the model in this situation
knows \textit{where} to look, but not \textit{what} to look for. In Figure
\ref{fig:incorrect_ideal_84}, the features activated are mostly in the correct
location on the ant head, and the rough texture is clearly visible, yet the
model predicts the incorrect class \textit{smooth}. Similarly in Figure
\ref{fig:incorrect_ideal_177}, the features activated are also mostly in the
correct location, yet the model predicts the incorrect class \textit{smooth}. In
this case, it may be due to the pose of the ant being slightly different from
the average pose. In the incorrectly classified images with non-expected
features, analysis shows that the model is unable to find \textit{where} to
look, and obtains feature information from other parts of the ant or the
background. Cases where the image was correctly classified using the
non-expected features can basically be seen as noise. In order to further
analyze this class, we should introduce some parameter such as model confidence
to examine further.

>>>>

\begin{abstract}
    Ant cuticle texture presumably provides some type of function, and therefore
    is useful to research for ecological applications and bioinspired designs.
    In this study, we employ statistical image texture analysis and deep machine
    learning methods to automatically group similar ant species based on
    morphological features. We establish a public database of ant cuticle images
    for research. We provide a comparative study of the performance of image
    texture classification and deep machine learning methods on ant cuticle
    images. Our results show that the deep learning methods are giving higher
    accuracy in recognizing ant cuticle textures. Our experiments also reveal
    that the deep learning networks designed for image texture performs better
    than the general deep learning networks.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Research has focused on the macrostructures and
internal chemical components that describe the functionality of the exoskeleton
\cite{gunderson_insect_1989}. More recent work is being done to understand the
functional aspects of external cuticle micro sculpturing
\cite{muthukrishnan_insect_2020, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires expert training.
Therefore, automated entomology has attracted biologists and computer
scientists, and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can take advantage of existing work in general
image analysis methods.

@@

There is a large variety of ant species and most species are diverse in terms of
size, shape, behaviors, and cuticle textures \colorbox{green}{[references]}.
Based on our initial observation of the cuticle sculpturing in ants, it appears
that some type of textural structures can be identified in images. Therefore,
this image dataset may be worth to be explored from the image texture analysis
point of view for the further knowledge discovery. Image texture analysis has
been used in image processing for many decades \colorbox{green}{[references]}. Due to its
diversity of images on scales, smoothness, coarseness, microtexture, and
macrotexture occurring in images, it is a difficult problem for image textural
interpreters. With a rapid development of deep machine learning, deep neural
networks have been frequently used in image texture analysis \colorbox{green}{[references]}.

Our contribution is to establish an ant image dataset which can be used to
explore the categorization of different ant cuticle textures using image texture
analysis methods. \colorbox{green}{(comment: we may need to say how the dataset
    is established)} Our experiment shows that the deep learning networks with image
texture embedded in the design give a better accuracy than the K-views
clustering method.

This paper is organized as below:  Section 2 reviews the sculpturing
identification in ants, and image texture analysis and insect classification.
Section 3 describes the dataset preparation and image texture analysis methods
used in our experiments. Section 4 gives the experimental results, visualization
of both raw and processed images and analysis. The conclusion then follows.

\section{Related work on sculpturing identification in entomology, image texture
  analysis, and insect classification}

% \subsection{Sculpturing identification}

Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}.

@@

There is often substantial overlap among terms, and closely related cuticle
patterns are likely to be functionally similar. For example, the definition for
“imbricate” is, “partly overlapping and appearing like shingles on a roof or
scales on a fish,” which is difficult to distinguish from the definition of
tesselate, “made up of squares like a chess board, either in sculpturing or in
color.” However, we note that the terms used to describe the cuticle patterns
are similar to the terms used to describe textural patterns in texture analysis.

% \subsection{Texture analysis}

Image texture analysis is an important part of image interpretation as many
objects appearing in the textural patterns either partially or completely inside
an image. This phenomenon appears frequently in many images we encountered in
our environment as almost everybody is using the iPhone to capture the photos.
Scientists are also studying their subjects depending on taking the images
remotely such as satellite images. With an abundant number of textural images
surrounding us, it is essential to have a reliable automatic tool for the
textural image interpretation.

Computer methods are developed for textural images based on the spatial
relationship of gray levels of pixels encoded in the texture
\cite{haralick_image_1985,hung_image_2019}. Those methods by de-facto follow the
steps used in the statistical pattern recognition
\cite{fukunaga_introduction_2013}. Therefore, textural features are extracted,
and the classification and segmentation methods are used for the interpretation
either using the supervised or unsupervised approaches. The interpretation
method is usually taken and modified from the traditional pattern recognition
methods. For example, the gray-level co-occurrence matrix (GLCM) and local
binary patterns (LBP) are two of methods for extracting features
\cite{haralick_textural_1973, goos_gray_2000}.

Similar to the traditional pattern recognition, classical image texture analysis
methods can be grouped into four categories: statistical, structural,
model-based, and transform-based methods \cite{bharati_image_2004}. Among all of
the categories, the statistical method is often used by employing the method
from the pattern recognition principle.  In doing so, an image texture is
represented as feature vector using feature extraction methods and then fed into
the algorithm. Markov random field models are also studied for textural image
interpretation \cite{hassner_use_1981, cross_markov_1983}. The transform-based
methods use some functions to decompose an image texture into a set of basic
feature images. Gabor filters and Wavelet expansions are two of the widely used
approaches \cite{bovik_multichannel_1990}.

@@

Other network architectures for end-to-end texture learning are also available.
For example, in the deep multiple-attribute-perceived network (MAP-Net)
\cite{zhai_deep_2019}, multiple perceptual attributes are progressively learned
in a mutually reinforced manner through multiple branches. In the deep
structure-revealed network (DSR-Net) \cite{zhai_deep_2020}, inherent structural
representation for a texture pattern is obtained by employing a primitive
capturing module to learn spatial primitives and a dependency learning module to
capture the dependency among the primitives. In \cite{mao_deep_2021}, a residual
pooling layer consisting of a residual encoding module and an aggregation module
is used to generate discriminative features of low dimensions. In
\cite{peeples_histogram_2021}, a histogram layer is designed to compute local
spatial distribution of CNN features. In \cite{chen_deep_2021}, an innovative
aggregation module is presented to exploit statistical self-similarity across
layers. All these architectures customize the standard CNN structure to
accomplish the characterization of certain spatial, visual, and statistical
traits unique to textural images.

@@

% \subsection{Insect classification}

Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, e.g. \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, e.g. \textit{ant}, \textit{fly}, and \textit{bee}
\cite{deng_imagenet_2009}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

\section{Ant cuticle image dataset}

% \subsection{Dataset preparation} % preparation generation

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. We also
refer to these ant head images, which describe the pose of the ant, as ant
cuticule images, due to the clear cuticle texture present on the head. Some
randomly selected images from each category are shown in Figures
\ref{fig:smooth-cuticle-texture} and \ref{fig:rough-cuticle-texture}.

To begin, a master spreadsheet was created with the 2,499 different ant species
to be identified for the primary dataset. The team was trained to identify
cuticle sculpturing through a process which consisted of one 45-minute
introductory lesson explaining the project and texture categories. Then, the
team was given a training set of photos to identify from the genus
\textit{Polyrhachis}. The sculpture identification protocol describes the two
primary categories: rough and smooth.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/rough_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:rough-cuticle-texture}
\end{figure}

@@

To collect the images, the assistants followed the taxonomy information
available in the master spreadsheet to the appropriate
AntWeb\footnote{\url{https://www.antweb.org/}} page. In many cases, there are
multiple ant head images of the same species, and occasionally there are
multiple image resolution available from a single image. To simplify the data
collection process, the assistants were instructed to download the first ant
head image of the species being identified in the highest resolution possible.
Each image was named with an identifier that corresponds with the row number in
the master spreadsheet. The same ant head images that were downloaded in the
data collection phase were the same ones used in the sculpture identification
protocol. Ant species which did not have any images of the head were excluded
from the dataset. Additionally, ant species which only had a head image of a
queen ant were excluded from the dataset.

@@

The completed dataset of contains 2,499 images. 1072 samples of rough textured
ant cuticle textures comprise 43\% of the dataset. The remaining 1427 samples of
smooth textured ant cuticle textures comprise 57\% of the dataset. Due to the
variety of the ant head image attributes, we apply simple preprocessing before
the images are used in our model. We want the images to have a uniform size for
simplicity in our classification process. Since the ant head images are
typically centered in the image, we apply a center crop to each image to create
a square image of the same size. Once the image is square, we resize each image
to a fixed size of 256x256 pixels. We leave other discrepancies in the images
untouched.

\section{Methods for comparative analysis and experimental setup}

% \subsection{Classical texture analysis methods}

@@

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 10-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm. The
theta $\theta$ parameters in the Gabor filters used are multiples of $\pi/10$
radians starting from $0$ to $\pi$ radians. The Gabor filter lambda $\lambda$
parameter is equal to $\pi$.

% \subsection{Deep learning models used for the experiments}

Our first deep learning model is \textit{visual geometry group} (VGG), a
convolutional neural network that takes advantage of very small convolutional
filters in a deep network architecture \cite{simonyan_very_2015}. We compare
four architectures of VGG: VGG11, VGG13, VGG16, and VGG19. The primary
difference between the architectures is the number of layers in each model. Our
second model is \textit{residual network} (ResNet), a deep network architecture
that includes shortcut connections between layers (residual connections)
\cite{he_deep_2015}. We compare three architectures of ResNet: ResNet18,
ResNet50, and ResNet101. Again, the primary difference between the architectures
is the number of layers in each model.

% \subsection{Deep texture analysis method used for the experiments}

@@

To handle the imbalance of the dataset, we apply undersampling for each class for
the training dataset. By using random stratified sampling, we construct a
training set with 800 images per class. The remaining images are randomly split
between test and validation, which turns out to roughly a 60\%/20\%/20\% train,
test, and validation data split. With 272 rough samples and 627 smooth samples
left over after the stratified split, the test dataset has roughly 136 rough
samples and 313 smooth samples. Since these leftover samples are split with code
by 50\% there will be some rounding variance and therefore the test dataset
built at run-time will not always have exactly the same number of samples.

% \subsection{Evaluation}

% \subsection{Environment}

@@

\section{Results and analysis}

% \subsection{Experimental results}

To begin, we share the results on the each algorithm specified in the
methodology section using the evaluation metrics described. The results are
shown in Table \ref{tab:results}. It should be noted that due to the class
imbalance in the dataset, the F1 score is the preferable metric to the accuracy.
% Therefore, the results have been sorted by F1 score in descending order.

@@

% \subsection{Clustering segmentation}

% \subsection{t-SNE Visualization}

@@

In this section, we provide visualization of the fine-tuned ResNet101 model and
the randomly initialized ResNet101 model using t-SNE dimensionality reduction.
First, we run the dataset preprocessing method and initialize both models. Then,
both models are trained according to the training parameters and the state of
each model is saved. To visualize the deep extracted features, we modify each
model to obtain the embeddings of the second to last layer. Then, we use the
t-SNE algorithm to reduce the dimensionality of the embeddings to 2 dimensions.
We plot side-by-side the ground truth and predicted labels for each model.
Figure \ref{fig:resnet101_tsne} shows the results of the trained randomly
initialized model and Figure \ref{fig:fresnet101_tsne} shows the results of the
fine-tuned model. Based on the visual results of the t-SNE visualization, we can see that the
fine-tuned model learned a stronger separation of the two classes, which
reinforces the results that the fine-tuned model received a higher average
accuracy.

@@

% \subsection{GradCAM visualization}

Next, we provide some visual analysis of some correctly and
incorrectly classified images using GradCAM. We provide two categories and two
subcategories in our analysis. The two categories are correct and incorrect
classification. Regardless of the feature activation map, the correctly
classified images have the same predicted label as the ground truth, and
incorrectly classified images have different predicted labels. The two
subcategories are expected and non-expected feature activation. In the expected
case, the features that are used to compute the classification are the same as
the features used by the assistants in the sculpture identification process. In
general, the features used by the assistants are the textures of the cuticle on
the ant head. In the non-expected case, the features used to compute the
classification are not from the head, for example, from the background,
extraneous text, or the body of the ant. We used randomly selected images from
the dataset and the fine-tuned ResNet101 model to perform the analysis. We show
the GradCAM results in Figures \ref{fig:correct_ideal},
\ref{fig:correct_nonideal}, \ref{fig:incorrect_ideal}, and
\ref{fig:incorrect_nonideal}. The left image shows the preprocessed image input
to the model. The right image shows the GradCAM output based on the
classification. Four specimens were selected randomly from each category and
subcategory.

Correctly classified images which use the expected features show the ideal
performance of the model. Incorrectly classified images which use the expected
features should be further analyzed. In essence, the model in this situation
knows where to look, but not what to look for. In Figure
\ref{fig:incorrect_ideal_84}, the features activated are mostly in the correct
location on the ant head, and the rough texture is clearly visible, yet the
model predicts the incorrect class \textit{smooth}. Similarly in Figure
\ref{fig:incorrect_ideal_177}, the features activated are also mostly in the
correct location, yet the model predicts the incorrect class \textit{smooth}. In
this case, it may be due to the pose of the ant being slightly different from
the average pose. In the incorrectly classified images with non-expected
features, analysis shows that the model is unable to find \textit{where} to
look, and obtains feature information from other parts of the ant or the
background. Cases where the image was correctly classified using the
non-expected features can basically be seen as noise. In order to further
analyze this class, we should introduce some parameter such as model confidence
to examine further.
